---
title: "Exploratory Analysis"
output: html_document
---

## Project title

Music Metrics

### Authors

Isha Bhansali, Rajvansh Singh Dang, Ray Vardaro

### Date

Spring 2024

### Abstract

#No more than three sentences that summarize your project. Focus on the very most important aspects. (1) "Our main question is .... This question is important because .... To address the question, we will ...." (2) "We are concerned with ..., because .... To address this concern, we plan to ...." (3) "Consider that .... This is important because .... Accordingly, we plan to ...."

Our main question is - Do musical characteristics such as danceability, energy levels, and valence scores correlate to a song's popularity? This question is important because understanding the relationship between various musical aspects and track popularity might provide useful insights into listener tastes and behaviour. To address the question, we will analyze a dataset comprising information on the musical features and popularity ratings of numerous songs. We are concerned with identifying if certain musical genres yield more popular songs since it can shed light on the relationships between music genres and audience preferences, potentially altering business practices and artist strategies. To address this concern, we plan to conduct statistical analyses comparing the popularity of songs in various genres. Consider how certain musicians constantly receive greater popularity scores. This is important because it implies the presence of trends or patterns in the music industry that may inform commercial strategy or artistic decisions. Accordingly, we plan to investigate the features and attributes shared by artists with high popularity scores to better understand their impact on audience reaction.

### Keywords

Spotify Music, Song Danceability, Song Energy

### Introduction  

Our exploratory project is based on the Spotify Tracks Dataset by the user maharshipandya. In this dataset, over 114,000 songs were collected and filtered by unique factors such as song name, artist name, popularity, song danceability and energy levels. With these various factors, our group is able to establish questions that have correlating trends from the collected data.

•	Do musical characteristics such as danceability and song energy levels correlate to popularity?
•	Do high valence scores relate to a song’s popularity?
•	Do certain musical genres produce more popular songs? 
•	Which artists have the highest popularity scores and is there a common trend?

Our group is interested in finding the correlation between artists, genres, popularity, and musical characteristics. Through this data, we can find information like what effect factors such as danceability, valence, and energy have on their listeners. We can also easily track how both artists and genres can relate to song popularity. Why is this research important to explore?  Our group believes music is a form of art that can affect its listeners, which is important to explore. By determining how certain musical characteristics relate to popularity, we can understand what specific elements of music people enjoy. This data can also explore how music from both popular and unpopular artists can affect their audiences. 

### Related Work  

For our group project, we aim to find the connections within music by exploring the artists, genres, popularity, while comparing them to song danceability, valence and energy levels. While many know that Spotify uses algorithms to align listeners to artists and songs, these systems are based on the listeners’ libraries. In the CNBC article “How Spotify AI plans to know before you do what your brain wants to listen to”, Spotify states investing in machine learning to better fit audiences to their preferred music choices (Jockims. How Spotify AI plans to know before you do what your brain wants to listen to. April 14 2024.) In the Pfizer study “Why — and How — Music Moves Us”, music is described to improve both mood and memory, which is tied heavily to the listeners’ emotional attachment (Fink, Jennifer. Why — and How — Music Moves Us. 2024.). In the Medium article, “Listening to This Music Makes you Happy!”, the relation of high valence, danceability, and energy scores to happiness over multiple generations (Twaddle, Alan. Listening to this music will make you happy!. 20, Aug 2021.). This article also outlines that valence driven happiness can derive from danceability, energy, speechlessness, and even specific key usage. 

### The Dataset

#Where did you find the data? 

I found data through the Google Datasets search. The website name is huggingface.co. Here is the link -> https://huggingface.co/datasets/maharshipandya/spotify-tracks-dataset

#Who collected the data? 

Spotify (maharshipandya/spotify-tracks-dataset)

#How was the data collected or generated?  

The data was collected and cleaned using Spotify's Web API and Python (mentioned on the website)

#Why was the data collected?  

Mentioned on the website: 
1. Building a Recommendation System based on some user input or preference
2. Classification purposes based on audio features and available genres

#How many observations (rows) are in your data?  

114,000 rows in total.

#How many features (columns) are in the data?

20 columns 

#What, if any, ethical questions or questions of power do you need to consider when working with this data?  

1. Privacy concerns 
2. Bias in Popularity Metrics
3. Representation and Diversity
4. Impact on the Music Industry

#What are possible limitations or problems with this data?   (at least 200 words)

This dataset may contain several restrictions or concerns. The sampling bias comes first. In statistics, sample bias can occur, resulting in specific song genres or artists being overrepresented or underrepresented. This results in skewed or discrepant research findings. Furthermore, because this information is from a certain period, there may be instances where new songs have been published and are the most popular. This may result in bias. 
Another example is that everyone classifies songs differently based on genre. As a result, errors or inconsistencies may arise along with disparities between the research findings and the dataset. This also explains how we may have little context to improve our comprehension of the dataset and complete our research. We will have to either ignore it or find a way to resolve it. 
Because the dataset only captures music over time, old tracks may become popular or new tracks may become renowned, resulting in excessive variation in the music industry. This, too, cannot be included in the dataset. It would be preferable to use longitudinal data or to update the dataset based on recent information. In addition, as time passes, music evolves in terms of background music, and people experiment with different styles. As a result, one issue is that the dataset does not accurately reflect contemporary music business tastes. 
The last point is based on all of the reasons mentioned above, as well as the fact that the songs used in the dataset are only from the United States and do not contain global variety, as Spanish and Hindi music have recently gained popularity, resulting in dataset generalization. 

4. Implications

### Implications

#Assuming you answer your research questions, briefly describe the expected or possible implications for technologists, designers, and policymakers. (at least 150 words)

The possible implications for technologists, designers and policymakers are:
1.	Technologists – The findings from the study can be used to improve algorithm recommendations. It also gives us additional information that we may use to improve the algorithm's efficiency. Understanding the links between music selections and their popularity, genre, instruments played, and other factors can also assist the algorithm improve music discovery and recommend new options. 
2.	Designers - They might include the research findings into the user app interface. For example, put popular tracks first or include them in recommended albums. They can utilize this function to promote niche genres or styles that are underrepresented in mainstream media. 
3.	Policymakers – They can utilize this information to understand the music industry better. This could lead to stronger business regulation and efforts to support artists through licensing and copyright. They might also implement more data-driven strategies to support budding musicians by analyzing how people react to certain tracks.

### Limitations & Challenges
#What challenges or limitations might you need to address with your project idea more broadly? Briefly discuss. (at least 150 words)

There may be some challenges and limits with the dataset. One of them is the accuracy of the data extracted from the dataset. Any minute mistakes or omissions in recording the song's timing, or a lack of information on when it was released. Another worry is the ethical considerations around the data. We don't know if the information was given with consent or if there were any privacy or representation issues. 
Another problem is that this dataset is based on a specific timeframe, and we must match the dataset's insight to the study questions based on that timeframe. We can't apply the same findings to current music trends because they're so dynamic. The music industry undergoes numerous changes that are difficult to predict using the data we have. We'll need more recent data to do this. We also don't know how the dataset rubrics were calculated. 
For example, the popularity of songs we know is graded from 0 to 100, but we don't know how the figures were computed. These are some challenges and limitations we will have to address with our project. 

### Summary Information
Our group was focused on collecting approximately eight data points from the Spotify Tracks Dataset by the user maharshipandya. These eight data points include track genre, popularity, duration, danceability, energy, loudness, tempo, and track count. The summary information table has been ranked by average popularity in order to better identify trends of track genres within the Music dataset. 

data<- read.csv("dataset.csv")
summary_info <- list(data)
summary_info$num_observations <- nrow(data)
summary_info$unique_track_genres <- length(unique(data$track_count))
summary_info$most_popular_songs <- data %>%
  filter(popularity == max(popularity, na.rm = T)) %>%
  select(track_genre, artists)
summary_info$longest_song_duration <- data %>%
  filter(duration_ms == max(duration_ms, na.rm = T)) %>%
  select(track_name)
summary_info$danceability <- data %>%
  filter(danceability == max(danceability, na.rm =T))%>%
  select(track_genre, artists)
summary_info$energy <- data %>%
  filter(energy == max(energy , ra.rm= T)) %>%
  select(track_name, track_genre)
summary_info$loudness <- length(unique(data$loudness))
summary_info$tempo <- data %>%
  filter(tempo == max(tempo, na.rm =T)) %>%
  select(track_genre)
print(summary_info)

### Table
library(readr)
library(dplyr)
import_dataset<- read_csv("dataset.csv")
aggregate_table <- import_dataset %>% group_by( artists, track_name, popularity, danceability, valence) %>%
  summarise(Song_score = sum(popularity+valence+danceability), .groups = 'drop')

print(aggregate_table)

``````{r, echo = FALSE, code = readLines("aggregate_table")}
library(readr)
library(dplyr)
import_dataset<- read_csv("dataset.csv")
aggregate_table = import_dataset %>% group_by( artists, track_name, popularity, danceability, valence, energy) %>%
  summarise(Song_score = sum(popularity+valence+danceability+energy), .groups = 'drop')
View(aggregate_table)
``````

In the data table, the aggregate information has been sorted by Artist Name, Track Name, Song Popularity, Song Danceability, Song Valence and Song Energy. By combining the Song Popularity, Danceability, Valence, and Energy, the Total Song Score data can be collected.  

### Chart 1

``````{r, echo = FALSE, code = readLines("Firstchart.R")}
library(readr)
library(ggplot2) 

dataset <- read.csv("dataset.csv")

ggplot(dataset, aes(x = danceability, y = energy)) +
  geom_point() +
  labs(title = "Danceability vs. Energy",
       x = "Danceability", y = "Energy")
print (ggplot)
``````
# Purpose - The graph helps identify the relationship between the two variables Energy and Danceability. It helps us visualize the common perception that higher energy songs are like to have high danceability. Since both the variable are numbers using a scatter plot also helps understand the relationship much better

# Insights - We see from the graph that there seems to be a positive relationship between the two variables. Though we see a lot of data crowded between 0.25-0.75 of the dancebility we can assume they fall in the normal bell shaped curves. We may conclude that most of the songs have a good energy hence resulting in danceability. But we may also say that due to the song having high danceability people assume it to have high energy. Hence, we know they share a positive relationship but we cannot assume one cause the other since there are several factors that contribute into this relationship. 


### Chart 2

``````{r, echo = FALSE, code = readLines("Secondchart.R")}

# Load the necessary libraries
library(dplyr)
library(ggplot2)
library(readr)
library(tidyverse)
dataset <- read.csv("dataset.csv")

avg_popularity <- aggregate(popularity ~ track_genre, data = dataset, FUN = mean)

# New aggregation to sum up popularity if track_genre is the same
sum_popularity <- dataset %>%
  group_by(track_genre) %>%
  summarise(total_popularity = sum(popularity))

# Combine both aggregations
combined_popularity <- left_join(avg_popularity, sum_popularity, by = "track_genre")

# Step 4: Create the bar chart using ggplot2
ggplot(combined_popularity, aes(y = track_genre, x = popularity)) +
  geom_bar(stat = "identity", fill = "skyblue", width = 0.7) + # Adjust width for space between bars
  geom_text(aes(label = track_genre), position = position_dodge(width = 0.7), vjust = -0.2, color = "black", size = 3) +  # Add text labels for total popularity
  theme_minimal() +
  labs(title = "Average and Total Popularity by Genre",
       y = "Genre",
       x = "Popularity")
``````
#Purpose - We decided to use a bar graph for the depiction between genre and popularity as it helps in to comparing different groups or track changes over time. It explains the complex data and gives us an overview of what type of music is the most popular. 
#Insights - As we see from the following bar graph, the top six genres from most popular to the least are Pop-film, K-pop, chill, sad,grunge and Indian. The least popular genres are romance, Iranian.This helps us understand how people would mostly react to the various genres.

### Chart 3

``````{r, echo = FALSE, code = readLines("Thirdchart.R")}
install.packages("gridExtra")
library("gridExtra")
# Download the dataset and read it into R
dataset <- read.csv("dataset.csv")

# Select the columns for the audio features
audio_features <- dataset[, c("danceability", "energy", "speechiness", "acousticness", "instrumentalness", "liveness", "valence", "tempo")]

# Create histograms for each audio feature
histograms <- lapply(names(audio_features), function(feature) {
  ggplot(dataset, aes_string(x = feature)) +
    geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black", alpha = 0.7) +
    labs(title = paste("Histogram of", feature),
         x = feature,
         y = "Frequency") +
    theme_minimal()
})

# Arrange histograms in a grid
grid.arrange(grobs = histograms, ncol = 2)
``````
#Purpose -  The graphs help us analyze the distribution of various audio features within a datasheet and visualize them using histograms.We can use these graphs to infer the distribution and range of values for all the audios. It also helps give us a deeper analysis into the various categories and identify the patterns or any specific trends
# Insights - We see that overall as per dancebility, energy and valence all have high frequency in most of the songs. Whereas, speechiness, acousticness, instrumentalness, liveness all have a negitive relationship meaning that their frequency decreases as the songs become more popular. While onlt the tempo follows a normal bell curve where most of the songs fall in the 50% of the graph.


###References

References:
1.	 Jockims, Trevor Laurence. “How Spotify AI Plans to Know before You Do What Your Brain Wants to Listen To.” CNBC, CNBC, 18 Apr. 2024, www.cnbc.com/2024/04/14/how-spotify-ai-plans-to-know-whats-going-on-inside-your-head.html 
2.	Fink, Jennifer. “Why — and How — Music Moves Us” Pfizer, 2024, www.pfizer.com/news/articles/why_and_how_music_moves_us#:~:text=Music%20and%20Mood&text=The%20limbic%20system%2C%20which%20is,when%20our%20ears%20perceive%20music.&text=The%20chills%20you%20feel%20when,of%20pleasure%20and%20well%2Dbeing
3.	Twaddle, Alan. “Listening to This Music Will Make You Happy!” Medium, Medium, 20 Aug. 2021, https://www.medium.com/@twaddlac/listening-to-this-music-will-make-you-happy-751182d8c5cd 